---
title: "Untitled"
author: "Kelly Bodwin"
date: "8/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs}
library(tidyverse)
library(tidymodels)
library(lubridate)
library(modelr)
```

```{r read data}
quant_norm <- function(vec){
  
  quants <- rank(vec)/(length(vec) + 1)
  normies <- purrr::map_dbl(quants, qnorm)
  
  return(normies)
  
}


dat_orig <- read_csv(here::here("MLB_combined.csv")) 
```

```{r data prep}
mlb_hype <- read_csv("mlb_hype.csv")

mlb_hype <- mlb_hype %>%
  mutate(
    Month = paste0(Month, "-01"),
    date = ymd(Month)
  )


dat <- dat_orig %>%
  left_join(mlb_hype) %>%
  mutate(
    googlehits_orig = googlehits,
    googlehits = quant_norm(googlehits),
    mlb_searches_orig = mlb_searches,
    mlb_searches = quant_norm(mlb_searches),
    playoffs = factor(playoffs, 
                      levels = c("No Playoffs", "Lost WC", "Lost Quarter", "Lost Semi", "Lost WS", "Won WS"))
  ) 

dat %>%
  write.csv("Model_Data.csv")
# 
# dat_split <- initial_split(dat)
# 
# train_dat <- training(dat_split)
# test_dat <- testing(dat_split)
```

## Notes

Why might a team have higher google hits in a given month?

* *Market Size*:  More fans = more people googling.

* *Leaguewide Interest*:  More people follow baseball during the playoffs.

* *Prior Performance:*  A team that wins the WS will be followed more closely the next year.

* *Expectations*: A team that is considered likely to content is in the news more.

* *Recent Performance:* A team on a hot streak will be in the news.

* *Special Events:* A trade, an off-field scandal, etc.



Predict each month based on these things?

```{r}
dat %>%
  mutate(
    mlb_searches = quant_norm(mlb_searches)
  ) %>%
  filter(team %in% c("San Francisco Giants", "Chicago Cubs")) %>%
  ggplot(aes(x = date, y = googlehits, color = team)) +
  geom_line() +
  geom_line(aes(x = date, y = mlb_searches), color = "black")
```


```{r}
dat_overall <- dat %>%
  group_by(team) %>%
  summarize(
    ghits_overall = mean(googlehits),
    propTVMarketSize = propTVMarketSize[1]
  ) %>%
  ungroup()

dat_annual <- dat %>%
  group_by(team, year) %>%
  summarize(
    ghits_annual = mean(googlehits),
    mlb_annual = mean(mlb_searches),
    propTVMarketSize = propTVMarketSize[1],
    playoffs = playoffs[1],
    odds = odds[1],
    winpct = winpct[1],
    attendance = attendance[1],
    payroll = payroll[1]
  ) %>%
  ungroup()

dat_annual <- dat_overall %>%
  select(team, ghits_overall) %>%
  right_join(dat_annual)


dat <- dat_annual %>%
  select(team, year, ghits_annual, mlb_annual, ghits_overall) %>%
  right_join(dat)
```


```{r overall model}
fit_overall <- lm(ghits_overall ~ propTVMarketSize,
                      data = dat_overall)

summary(fit_overall)

dat_annual <- dat_annual %>%
  add_predictions(
    fit_overall,
    var = "pred_overall"
  ) %>%
  mutate(
    resid_annual = ghits_annual - pred_overall
  )


```


```{r annual model}
fit_annual <- lm(resid_annual ~ mlb_annual + 
                   lag(ghits_annual) + 
                   lag(playoffs) + lag(winpct) + lag(attendance) +
                   odds + payroll,
                      data = dat_annual) 

summary(fit_annual)

dat <- dat %>%
  add_predictions(
    fit_overall,
    var = "pred_overall"
  ) %>%
  add_predictions(
    fit_annual,
    var = "pred_resid_annual"
  ) %>%
  mutate(
    pred_annual = pred_overall + pred_resid_annual,
    resid = googlehits - pred_annual
  )
```


```{r monthly model}
fit_monthly <- lm(resid ~ mlb_searches + lag(googlehits) + 
                        phase*(winpct + attendance) + is_playoffs:playoffs,
                      data = dat) 

summary(fit_monthly)

dat <- dat %>%
  add_predictions(
    fit_monthly,
    var = "pred_resid"
  ) %>%
  mutate(
    pred = pred_overall + pred_resid_annual + pred_resid,
    resid_final = googlehits - pred
  )
```


```{r final resids}
dat %>%
  filter(team %in% c("San Francisco Giants", "Los Angeles Dodgers", "San Diego Padres")) %>%
  ggplot(aes(x = date, y = resid_final, color = team)) +
  geom_line()


dat %>%
  filter(team %in% c("San Francisco Giants")) %>%
  ggplot() +
  geom_line(aes(x = date, y = googlehits), color = "black") +
  geom_line(aes(x = date, y = pred), color = "red")

dat %>%
  filter(team %in% c("Chicago Cubs")) %>%
  ggplot() +
  geom_line(aes(x = date, y = googlehits), color = "black") +
  geom_line(aes(x = date, y = pred), color = "red")
```




$$\text{\{Google Hits\}}_i \sim \alpha_i \, \text{\{TV Market Size\}_i} + E_1 + \epsilon_1$$

where $\epsilon \sim N(0, \sigma^2)$

$$E_{1ij} \sim \gamma_{ij} ... + E_2$$

## Team-specific

```{r}
lm(resid_final ~ team, data = dat)
```



## Extra

```{r}
dat %>%
  filter(phase == "Playoffs") %>%
  lm(googlehits ~ propTVMarketSize+ playoffs + winpct, data = .) %>% 
  tidy() %>%
  ggplot(aes(x = fct_inorder(term), y = estimate)) +
  geom_col()


dat %>%
  filter(phase == "Season") %>%
  lm(googlehits ~ propTVMarketSize + attendance + odds + winpct, data = .) %>% 
  tidy() %>%
  ggplot(aes(x = fct_inorder(term), y = estimate)) +
  geom_col()
```

```{r}
by_phase <- dat %>%
  mutate(
    year = year(date)
  ) %>%
  drop_na() %>%
  group_by(team, year) %>%
  mutate(googlehits = scale(googlehits)) %>%
  group_by(team, year, phase) %>%
  summarize(hits = mean(googlehits)) %>%
  pivot_wider(names_from = phase, 
              values_from = hits) %>%
  ungroup() 

plot(by_phase$Playoffs, by_phase$Season)

by_phase %>%
  select(Opening, Playoffs, Season, `Off-Season`) %>%
  drop_na() %>%
  as.matrix() %>%
  cor()
```


